Exploring how changes in cause of death have narrowed the black-white life expectancy gap in the United States.

### 2017-02-23 Manuscript/Manuscript_COD_decomp.Rmd
updating ms

bw_dat.females <- bw_dat %>% filter(sex == "Female") %>%
  mutate(era1.diff = first.gap - second.gap,
         era2.diff = second.gap - third.gap,
         era3.diff = third.gap - fourth.gap,
         state.gdiff.order.1 = reorder(state, first.gap),
         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
         state.gdiff.order.2 = reorder(state, second.gap),
         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
         state.gdiff.order.3 = reorder(state, third.gap),
         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
  )

# cod.df <- data.frame(state = factor(), sex = factor(), era = character(), 
#                      COD = factor(), Contribution.to.change = numeric(), 
#                      Contrib.to.change.prop = numeric(), narrowed_gap = logical())
# 
# age.df <- data.frame(state = factor(), sex = factor(), era = character(), 
#                      age_minbin = numeric(), Contribution.to.change = numeric(), 
#                      Contrib.to.change.prop = numeric(), narrowed_gap = logical())
# 
# years <- c(1969, 1983, 1993, 2013)
# 
# for (state.i in levels(BlackWhite_results$state)) {
#   for (sex.i in c("Male", "Female")) {
#     for (year.i in 1:3) {
#       
#       table1 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(COD)
#       table2 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(COD)
#       
#       table3 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(age_minbin)
#       table4 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(age_minbin)
#       
#       temp1 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
#                           contribution.to.gap.change(type.of.decomp = "COD", 
#                                                      decomp.table1 = table1, 
#                                                      decomp.table2 = table2, 
#                                                      type.of.data = "results"))
#       
#       temp2 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
#                           contribution.to.gap.change(type.of.decomp = "Age", 
#                                                      decomp.table1 = table3, 
#                                                      decomp.table2 = table4,
#                                                      type.of.data = "results"))
#       
#       cod.df <- rbind(cod.df, temp1)
#       age.df <- rbind(age.df, temp2)
#     }
#   }
# }

```{r more_figure_prep, echo = F}
cod.df <- cod_change_results %>% select(state, sex, Cause.of.death, Cx_era1_diff_mean, Cx_era2_diff_mean, Cx_era3_diff_mean)
cod.df <- tidyr::gather(cod.df, Cx_era1_diff_mean, Cx_era2_diff_mean, Cx_era3_diff_mean, key = "era", value = "Cx_diff_mean")
ncols <- str_split_fixed(as.character(cod.df$era), "_", 4)
cod.df$era <- as.factor(ncols[ , 2])

@@

@@

In 1969, the life expectancy gap ranged from `r round(sum1[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `. Injury was the largest contributor towards narrowing the gap during this time period, as both race groups experienced decreased rates of injury-related mortality, but blacks had larger rates decreases, especially in states with relatively higher rates in 1969 (**S3-S14 Figs**). A few states, including Minnesota and Nebraska, had relatively low rates of injuries among blacks in 1969 and smaller gaps. The residual category also contributed to narrower gaps in most states, but generally made a smaller contribution than injuries. In New York, the residual category made a large contribution to narrowing the gap, where much of this decrease was likely due to decreased infant mortality among blacks (**S15 Fig**). Communicable diseases contributed modestly to narrowing the gap, except in Arizona and Minnesota. Generally, cardiovascular disease contributed towards increasing the gap, and reflects slower declines for blacks in cardiovascular mortality than for whites. Cancer contributed towards increasing the gap in most states, and largely reflects increased cancer-related mortality for both races during this period, with blacks undergoing steeper changes. In most states, non-communicable diseases contributed a small amount towards increasing the gap, although it decreased the gap among four states, most notably in New York.

@@

The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC (**Fig 3**). In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` years [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum3[dim(sum3)[1], ]$state `. By 2013, the gap ranged from `r round(sum4[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` years [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum4[dim(sum4)[1], ]$state `. During this time period all of the causes contributed to narrowing the gap for most states. In Wisconsin and the District of Columbia, changes to CVD mortality counteracted the gap narrowing (reflecting slower decreases to cardiovascular mortality among blacks compared with whites), while in Arizona changes to communicable disease counteracted the narrowing.

@@

males.era1 <- ggplotly(
  ggplot(subset(cod.df.males, era == "era1"), aes(y = state.gdiff.num.1, x = new.start.1)) +
    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.1, fill = Cause.of.death), 
              color = "white") +
    geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
        geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) + 
    theme_minimal() + 
    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
    scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

males.era1
```

@@

![Difference in life expectancy between black in white males, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Men.bmp)

@@

![Difference in life expectancy between black in white females, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Women.bmp)

@@

```{r ny_1969_1983_age_cod_decomp, echo = F, fig.width = 9}
ny.dat <- age_cod_results %>% filter(state == "New York" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

@@

**Pennsylvania**
```{r pa_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
pa.dat <- age_cod_results %>% filter(state == "Pennsylvania" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

@@

**Connecticut**
```{r ct_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
ct.dat <- age_cod_results %>% filter(state == "Connecticut" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

@@

**Mississippi**
```{r ms_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
ms.dat <- age_cod_results %>% filter(state == "Mississippi" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

@@

**Florida**
```{r fl_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
fl.dat <- age_cod_results %>% filter(state == "Florida" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

@@

>>>>

bw_dat.females <- bw_dat %>% filter(sex == "Female") %>%
  mutate(era1.diff = first.gap - second.gap,
         era2.diff = second.gap - third.gap,
         era3.diff = third.gap - fourth.gap,
         state.gdiff.order.1 = reorder(state, first.gap),
         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
         state.gdiff.order.2 = reorder(state, second.gap),
         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
         state.gdiff.order.3 = reorder(state, third.gap),
         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
  )
```

```{r more_figure_prep, echo = F}
widths <- cod_change_results %>% mutate(tot_Cx_era1_CI_width = tot_Cx_era1_diff_ucl - tot_Cx_era1_diff_lcl,
                                        tot_Cx_era2_CI_width = tot_Cx_era2_diff_ucl - tot_Cx_era2_diff_lcl,
                                        tot_Cx_era3_CI_width = tot_Cx_era3_diff_ucl - tot_Cx_era3_diff_lcl) %>%
  filter(Cause.of.death == "Injuries") %>%
  select(state, sex, tot_Cx_era1_CI_width, tot_Cx_era2_CI_width, tot_Cx_era3_CI_width)

widths <- tidyr::gather(widths, tot_Cx_era1_CI_width, tot_Cx_era2_CI_width, tot_Cx_era3_CI_width, key = "era", value = "tot_Cx_CI_width")
ncols <- str_split_fixed(as.character(widths$era), "_", 5)
widths$era <- as.factor(ncols[ , 3])

cod.df <- cod_change_results %>% select(state, sex, Cause.of.death, Cx_era1_diff_mean, Cx_era2_diff_mean, Cx_era3_diff_mean)
cod.df <- tidyr::gather(cod.df, Cx_era1_diff_mean, Cx_era2_diff_mean, Cx_era3_diff_mean, key = "era", value = "Cx_diff_mean")
ncols <- str_split_fixed(as.character(cod.df$era), "_", 4)
cod.df$era <- as.factor(ncols[ , 2])

@@

cod.df <- merge(cod.df, widths, by = c("state", "sex", "era"))

@@

In 1969, the life expectancy gap ranged from `r round(sum1[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap_mean, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `. Injury was the largest contributor towards narrowing the gap during this time period, as both race groups experienced decreased rates of injury-related mortality, but blacks had larger rates decreases, especially in states with relatively higher rates in 1969 (**S3-S14 Figs**). A few states, including Minnesota and Nebraska, had relatively low rates of injuries among blacks in 1969 and smaller gaps. The residual category also contributed to narrower gaps in most states, but generally made a smaller contribution than injuries. In New York, the residual category made a large contribution to narrowing the gap, where much of this decrease was likely due to decreased infant mortality among blacks (**S15 Fig**). Communicable diseases contributed modestly to narrowing the gap, except in Arizona and Minnesota. Generally, cardiovascular disease contributed towards increasing the gap, and reflects slower declines for blacks in cardiovascular mortality than for whites. Cancer contributed towards increasing the gap in most states, and largely reflects increased cancer-related mortality for both races during this period, with blacks undergoing steeper changes. In most states, non-communicable diseases contributed a small amount towards increasing the gap, although it decreased the gap among four states, most notably in New York.

@@

The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC (**Fig 3**). In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum3[dim(sum3)[1], ]$state `. By 2013, the gap ranged from `r round(sum4[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` years [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum4[dim(sum4)[1], ]$state `. During this time period all of the causes contributed to narrowing the gap for most states. In Wisconsin and the District of Columbia, changes to CVD mortality counteracted the gap narrowing (reflecting slower decreases to cardiovascular mortality among blacks compared with whites), while in Arizona changes to communicable disease counteracted the narrowing.

@@

males.era1 <- ggplotly(
  ggplot(subset(cod.df.males, era == "era1"), aes(y = state.gdiff.num.1, x = new.start.1)) +
    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.1, fill = Cause.of.death), 
              color = "white") + # + scale_alpha(range = c(0.64, 4.04) 
    geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
        geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) + 
    theme_minimal() + 
    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
    scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

males.era1

plot <-   ggplot(subset(cod.df.males, era == "era1"), aes(y = state.gdiff.num.1, x = new.start.1)) +
    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.1, fill = Cause.of.death), #, alpha = tot_Cx_CI_width
              color = "white") + # + scale_alpha(range = c(0.64, 4.04) 
    geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
        geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) + 
    theme_minimal() + 
    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
    scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)
```

@@

![Difference in life expectancy between black and white males, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Men.bmp)

@@

![Difference in life expectancy between black and white females, United States, 1969-2013.](/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Manuscript/export.graph.Women.bmp)

@@

```{r ny_1969_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
ny.dat <- age_cod_results %>% filter(state == "New York" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

@@

**Pennsylvania**
```{r pa_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
pa.dat <- age_cod_results %>% filter(state == "Pennsylvania" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

@@

**Connecticut**
```{r ct_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
ct.dat <- age_cod_results %>% filter(state == "Connecticut" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

@@

**Mississippi**
```{r ms_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
ms.dat <- age_cod_results %>% filter(state == "Mississippi" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

@@

**Florida**
```{r fl_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
fl.dat <- age_cod_results %>% filter(state == "Florida" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

@@

**S2 Table: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**

```{r}
cod_change_results2 <- data.frame(cod_change_results[c(3, 40, 41)], apply(cod_change_results[4:39], 2, function(x) x*(-1)))

cod_change_results2 <- cod_change_results2 %>% 
  mutate(total_era1_diff = paste0(round(tot_Cx_era1_diff_mean, 2), " [", round(tot_Cx_era1_diff_lcl, 2), ",", round(tot_Cx_era1_diff_ucl, 2), "]"),
         total_era2_diff = paste0(round(tot_Cx_era2_diff_mean, 2), " [", round(tot_Cx_era2_diff_lcl, 2), ",", round(tot_Cx_era2_diff_ucl, 2), "]"),
         total_era3_diff = paste0(round(tot_Cx_era1_diff_mean, 2), " [", round(tot_Cx_era3_diff_lcl, 2), ",", round(tot_Cx_era3_diff_ucl, 2), "]"),
         contribution_era1_diff = paste0(round(Cx_era1_diff_mean, 2)," [", round(Cx_era1_diff_lcl, 2), ",", round(Cx_era1_diff_ucl, 2), "]"),
         contribution_era2_diff = paste0(round(Cx_era2_diff_mean, 2)," [", round(Cx_era2_diff_lcl, 2), ",", round(Cx_era2_diff_ucl, 2), "]"),
         contribution_era3_diff = paste0(round(Cx_era3_diff_mean, 2)," [", round(Cx_era3_diff_lcl, 2), ",", round(Cx_era3_diff_ucl, 2), "]")
         ) %>%
  select(sex, state, Cause.of.death, total_era1_diff, contribution_era1_diff, total_era2_diff, contribution_era2_diff, total_era3_diff, contribution_era3_diff) %>%
  rename(Sex = sex, State = state, 
         `Change from 1969 to 1983, years` = total_era1_diff,
         `Change from 1983 to 1993, years` = total_era2_diff,
         `Change from 1993 to 2013, years` = total_era3_diff,
         `Contribution to gap change ('69-'83), years` = contribution_era1_diff,
         `Contribution to gap change ('83-'93), years` = contribution_era2_diff,
         `Contribution to gap change ('93-'03), years` = contribution_era3_diff)

data.table::data.table(cod_change_results2 %>% filter(Sex == "Male"))
```

**S3 Table: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 1983 by US state**

```{r}
data.table::data.table(cod_change_results2 %>% filter(Sex == "Female"))
```
